---
title: "LA Street Slopes"
output:
  html_notebook: default
  html_document: default
  toc: yes
---
### Project Background
For the LADOT Play Streets Program, I was asked if we had a layer that had the following information: 

1. Street Classification
2. Street Width
3. Street Slope

The street classification system is part of the street centerline file, maintained by the Los Angeles Department of Public Works Bureau of Engineering and available at the City's GeoHub [here](http://geohub.lacity.org/datasets/d3cd48afaacd4913b923fd98c6591276_36). Street width is contained in a slightly different version of the same centerline file, maintained by the Bureau of Street Services, and available as part of the 'Street Pavement Condition' layer on the City's GeoHub [here](http://geohub.lacity.org/datasets/2d945910272548a6ae0b57126000669a_0).

However, there is not readily available layer with street slope already calculated, so I decided to create one myself, using [this blog post](http://www.datapointed.net/2009/11/the-steeps-of-san-francisco/) as a guide. The author, Stephen Von Worley, combined the [National Elevation Dataset's](https://nationalmap.gov/elevation.html) 1/3-second data with the [OpenStreetMap](http://www.openstreetmap.org/) grid to produce his result. For this project, I combined the same elevation data with the two layers I mentioned above. Another option in R that I could have pursued included utiling the R package `elevatr`, which includes querys for elevation data for a bunch of different APIs. Most of the APIs don't provide access to Digital Elevation Models, which is the most precise elevation data, and definitely necessary for this project. Mapzen, though it does provide API access to raster layers and DEM-based querying API, it has rate limits, so in this case it was just easier to download the DEM raste data and query from there.

Before starting this project, I went ahead and downloaded the two street centerline shapefiles as well as the "USGS NED 1/3 arc-second 2013 1 x 1 degree ArcGrid" elevation data. For the City of Los Angeles, I needed to download two data elevation tiles: [n35w119](https://www.sciencebase.gov/catalog/item/581d2227e4b08da350d54373) (Northern LA), and [n34w119](https://www.sciencebase.gov/catalog/item/581d2217e4b08da350d541c7) (Southern LA). Also, in order to calculate the slope for each street segment, my plan was to calculate the elevation at each end point of the street segment. I could generate an endpoint file from the centerline network, but there is already one available on the GeoHub [here](http://geohub.lacity.org/datasets/0372aa1fb42a4e29adb9caadcfb210bb_9). This intersections layer has IDs that can be easily joined to the street centerline file for the slope calculation.

### Project Tools

* `sf` support for simple features; [here](http://strimas.com/r/tidy-sf/) is a good overview of the package

* `raster` For the analysis of the elevation data

* `tidyverse` Collection of R packages for datascience, including tidyr and dplyr

* `leaflet` for viewing our output on a web map 

```{r}
# load libraries
library(sf)
library(tidyverse)
library(leaflet)

# import street data
bss <- st_read("Data/Street_Pavement_Condition.shp",quiet=TRUE)
boe <- st_read("C:/Users/dotcid034/Documents/NonGitHubData/boe/Streets_Centerline.shp", quiet=TRUE)
int <- st_read("Data/Intersections.shp", quiet=TRUE)
```
### Explore the Data

__BOE Centerline__
```{r}
glimpse(boe)
```
__BSS Centerline / Street Pavement Condition__
```{r}
glimpse(bss)
```
### Clean / Prepare Intersection & Centerline Files
For the centerline files, both have quite a bit of unneeded information that we can trim out when we do the merge. Since the BSS version of the centerline file contains the street width and the BOE version contains the street classification, we need to merge them (and trim out any other unneeded columns) into one street centerline file. The BSS centerline file is slightly more restrictive -- only containing streets that are maintained by the City of Los Angeles -- so it will have fewer streets compared to the BOE one. I'll merge the BOE file with that one. 

Tip: Once I started bringing in the raster data, I began getting an error while runing the `select()` function. I found out from [this blog post](https://remi-daigle.github.io/2017-CHONe-Data/dataviz.nb.html#) that the error is a result of several packages having a `select()` function (in this case the function from the `raster` package is superceding the `dplyr` one). To solve the mixup, just make sure to explicitly designate which package the `select()` function is from using the syntax `package::function()` when it is not clear.
```{r}
# trim bss centerline
bss <- bss %>%
  dplyr::select(SECT_ID, PRIME, FROM_, TO_, SURF1, ZLENGTH, ZWIDTH, geometry) %>%
  rename(sect.id = 'SECT_ID',
         street = 'PRIME',
         from = 'FROM_',
         to = 'TO_',
         surface = 'SURF1',
         length.ft = 'ZLENGTH',
         width.ft = 'ZWIDTH',
         geom = 'geometry') %>%
  mutate(sect.id = as.double(sect.id))

# trim boe centerline
boe <- boe %>%
  dplyr::select(ASSETID, INT_ID_FRO, INT_ID_TO, SECT_ID, OLD_STREET, Street_Des) %>%
  rename(asset.id = 'ASSETID',
         fromint.id = 'INT_ID_FRO',
         toint.id = 'INT_ID_TO',
         sect.id = 'SECT_ID',
         old.desig = 'OLD_STREET',
         new.desig = 'Street_Des') %>%
  mutate(sect.id = as.double(sect.id)) %>%
  st_set_geometry(NULL)

# left join bss <- boe
streets <- bss %>%
  inner_join(boe, by='sect.id')
```
Let's take a peek at the merged dataframe:
```{r}
glimpse(streets)
```
For the intersection layer, the I'm just going to trim away unneeded columns. First, a glimpse:
```{r}
glimpse(int)
```
Now I'll trim away the unneeded columns:
```{r}
int <- int %>%
  dplyr::select(ASSETID, CL_NODE_ID, X, Y, LAT, LON, geometry) %>%
  rename(asset.id = 'ASSETID',
         clnode.id = 'CL_NODE_ID',
         x = 'X',
         y = 'Y',
         lat = 'LAT',
         lon = 'LON',
         geom = 'geometry')
```
### Extract Elevation Data from Raster Layer
In order to calculate the slope of each street segment, I first need to calculate the elevation at each of these street segment endpoints. In order to get the elevation at each of the endpoints (intersections), I am going to extract this elevation data from the raster elevation image that I pulled from the National Elevation Dataset.

But, before I even get there, I need to load the two raster data sets (one for the northern part of LA, one for the sourthern part of LA) and merge them using the `raster` package. We can then view the result:

```{r}
library(raster)

# import elevation data
la_n <- raster("C:/Users/dotcid034/Documents/NonGitHubData/n35w119/grdn35w119_13")
la_s <- raster("C:/Users/dotcid034/Documents/NonGitHubData/n34w119/grdn34w119_13")

# merge the raster files
la_raster <- merge(la_n, la_s)

# view the resulting pot
plot(la_raster)
```

-look at documentation for interpreting the extraction
- because the latest version of raster was before the release of th `sf` package, we are going to need to do some workarounds.
```{r}
int.df <- int %>%
  select(asset.id, lat, lon)

values <- extract(la_raster, int)
```

### Fun Aside: What are the steepest streets in LA?
-include a google streetview link at the locations

### Display the Map
```{r}
# map <- leaflet() %>%
#   addProviderTiles(providers$Stamen.TonerLite) %>%
#   addPolylines(
#     color = '#0E016F',
#     weight = 1,
#     opacity = 1,
#     data = streets
#   )
# 
# map
```
